---
title: "Report - diffPrivML"
author: "Terence Tam, Yifan Gong"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Description and problem statement

Machine learning algorithms collectively strive to extract population information and learn general patterns from data sets. The output of machine learning algorithms are generally trained models, where raw data points are distiled to far more compact forms that capture the essence of the origninal data. It is a type of summary aggregation in short that is no different than COUNT and SUM.  


The release of these trained machine learning models, however, comes with a cost--individual data privacy. Oftentimes, machine learning models inadvertently give away information on *individual* data points used in the training data, even though these ML models are meant to release *population* information only.  


To analyze and potentially solve this problem, one unified framework proposed by Cynthia Dwork's paper is the $\epsilon$-differential private mechanism. This is a model that quantifies the degrees of privacy loss for a given data release system, and a tolerance level as a parameter--$\epsilon$. A lot of theoretical research has been done on different ML models with regards to differential privacy, such as these research [here](https://arxiv.org/pdf/1412.7584.pdf), but not enough efforts are being done to apply these in practice.

Therefore, we developed a R package, **diffPrivML**, that aims to put these differential private enhanced ML algorithms in the hands of data practitioners. Just like how a unified framework is developed to measure differential privacy, **diffPrivML** is a precursor in standardizing a library of ML algorithms and workflows that have differential privacy baked in. End users do not need to worry about *HOW* to perturb the data to release their models; the ML algorithm itself makes the guarantee.

## Overview of package

The **diffPrivML** package consists of a library of classifiers implemented as S3 classes, each with differential privacy supported.

* DPNaiveBayesClassifier 
* DPRandomDecisionTreeClassifier
* DPLogisticRegressionClassifier 

For example, to fit a model of **DPNaiveBayesClassifier**, the privacy budget, epsilon, can be specified as part of the input like so.

```{r, eval=FALSE}
naiveBayesDP <- DPNaiveBayesClassifier(y = classLabels, x = predictors, epsilon = 0.05)
```

The model is fitted with the input training data when the classifier object gets instantiated.

Each classifier in **diffPrivML** also has all the standard ML methods such as **summary()** and **predict()**. So far, all the ML algorithms in this package uses output perturbation methods for enforcing differential privacy.  

All unit tests are in **/tests** directory.

One slight deviation from the original project proposal to also support neural network based classifier is that it is not straightforward to ship a default DF enhanced neural network classifier out of the box. Neural networks can be higly customized to first construct the neural network configuration (this is even before any training is done), and so the differential privacy parameters have to be customized as well based on the network constructed.

## Use cases

### Random Decision Tree Ensemble Classifier

First, let's invoke the caret library, just for the sake that it can cleanly partitiotn training vs test data for us.

```{r}
library(caret)
library(diffPrivML)
set.seed(50) # for reproducibility
```

Then, let's use the categorical health dataset **BreastCancer** preloaded in the diffPrivML package. We picked health records as that is the quintessential data that needs privatization!

```{r}
data(BreastCancer)
head(BreastCancer)
```

Now let's actually split the BreastCancer data into training set and test set, and fit the classifier model with the training data.

```{r}
Y <- BreastCancer[, 11]
X <- BreastCancer[, 2:10]

trainingIndices <-createDataPartition(Y, p = 0.70, list = FALSE)
trainingSetX <- X[trainingIndices, ]
trainingSetY <- Y[trainingIndices]
testSetX <- X[-trainingIndices, ]
testSetY <- Y[-trainingIndices]

rdtDP <- DPRandomDecisionTreeClassifier(Y = trainingSetY, X = trainingSetX,
                                        height = 4, numTrees = 5, epsilon = 1.0)
summary(rdtDP)
```

The summary() method will give all the parameters associated with the trained model. Again, end user can tweak parameters like depth of trees (height) and number of trees in the ensemble (numTrees) to boost the model performance.

For predicting the test set, we then run predict() on the trained model.

```{r}
predictions <- predict(rdtDP, testSetX)
accuracy <- sum(predictions == testSetY) / length(testSetY)
paste("The test set accuracy rate is", accuracy * 100, "%")
```

Note this classifier is only supported for categorical data. DPNaiveBayesClassifier, on the other hand, support mixed categorical / numerical attributes.

One thing end users of this package have to keep in mind is the tradeoff between how much to reveal in their trained model (privacy) vs how effective the classification is (accuracy). Below, we graphed the accuracy of the above dataset with various epsilon (privacy budget) to show this decay.

```{r}
eps <- c(5, 1, 0.7, 0.4, 0.01, 0.0002)

accuracy <- sapply(eps, function(epsilon) {
  rdtDP <- DPRandomDecisionTreeClassifier(Y = trainingSetY, X = trainingSetX, height = 3, 
                                          numTrees = 6, epsilon = epsilon)
  predictions <- predict(rdtDP, testSetX)
  sum(predictions == testSetY) * 100 / length(testSetY)
})

df <- data.frame(epsilon = eps,
                accuracy = accuracy)
ggplot(data=df, aes(x=epsilon, y=accuracy, group=1)) +
  geom_line() +
  geom_point() +
  scale_size_area()
```


### Naive Bayes Classifier

Just like the previous classifier, the same is true for **DPNaiveBayesClassifier** where a dataset is used as input.
```{r}
data(iris)
y <- iris[, 5]
x <- iris[, 1:4]
trainingIndices <-createDataPartition(y, p = 0.70, list = FALSE)
trainingSetX <- x[trainingIndices, ]
trainingSetY <- y[trainingIndices]
testSetX <- x[-trainingIndices, ]
testSetY <- y[-trainingIndices]

```

Then invoke the **DPNaiveBayesClassifier**, and also get the summary of the model fitted.

```{r}
naiveBayesDP <- DPNaiveBayesClassifier(trainingSetY, trainingSetX, epsilon = 0.7)
summary(naiveBayesDP)
```
Then try to predict the test data set like so.

```{r}
predictions <- predict(naiveBayesDP, testSetX)
accuracy <- sum(predictions == testSetY) / length(testSetY)
paste("The test set accuracy rate is", accuracy * 100, "%")
```

Trying on different epsilons.

```{r}
eps <- c(1, 0.7, 0.4, 0.1, 0.05)

accuracy <- sapply(eps, function(epsilon) {
  naiveBayesDP <- DPNaiveBayesClassifier(trainingSetY, trainingSetX, epsilon = epsilon)
  predictions <- predict(naiveBayesDP, testSetX)
  sum(predictions == testSetY) * 100 / length(testSetY)
})

df <- data.frame(epsilon = eps,
                accuracy = accuracy)
ggplot(data=df, aes(x=epsilon, y=accuracy, group=1)) +
  geom_line() +
  geom_point() +
  scale_size_area()
```

## Synthetic Data - Proof of differential privacy

It is trivial to capture what kind of privacy data loss a summary query would leak in a database system. For example, a pair of queries like **SELECT SUM(salary)** and **SELECT SUM(salary) WHERE name != ‘Terence Tam’** would do the trick. An adversary just observe the difference between the two results and be able to deduce the individual salary of "Terence Tam".

However, when it comes to the output of a machine learning model, it is not so straightforward as the "summary" returned is rather complex. In addition, understanding in the context of privacy loss the difference between how two training data sets that differs only by one datapoint can lead to two differently trained models is also nontrivial.

To be able to demonstrate this, I will use the **DPRandomDecisionTreeClassifier** again in this package. The advantage of decision tree is that it can be served as a good visualization of the output space created by the model. 

Decision trees are built where each predictor is a node, and each branch is one of the level of the predictor.

Suppose we have some synthetic data set that looks like this

```{r}
A <- c('0','1','1')
B <- c('1','0','0')
class <- c("Y","N","N")
syntheticData <- as.data.frame(cbind(A, B, class))
head(syntheticData)
```

A and B are predictors and class is the label for the data points. This is an extreme case where the whole training dataset only have three data points. Notice how the data points are very orthogonal and mutually exclusive, where class == 'Y' only when B == '1'. Therefore, in this toy example, knowing exactly the output of the classifier would definitely give inference on what data points are used in the training set, leading to privacy loss.

Suppose we run a non-private random decision tree classifier with the three data points as input like so.

```{r}
rdtDP <- DPRandomDecisionTreeClassifier(Y = syntheticData[,3], X = syntheticData[,1:2], height = 2, numTrees = 1)
```

When epsilon is not specified in our DP classifier, that will default to non-private algorithm (in a sense, that means epsilon is infinite). Also here, explicitly setting numTree = 1 so we only have 1 tree in our classifer for sake of discussion.



Note the RDT tree is unpruned. Now suppose we train the classifer 100 times, it is with certainty that with the data point, the predicted output is the same.

```{r}
experiment <- function(epsilon){
  rdtDP <- DPRandomDecisionTreeClassifier(Y = syntheticData[,3], X = syntheticData[,1:2], 
                                          height = 2, numTrees = 1, epsilon = epsilon)
  predict(rdtDP, syntheticData[1,1:2])
}
table(replicate(100, experiment(NULL)))
```

Now let's suppose we are limiting privacy budget, and then try compare the two models.

With data point

```{r}
table(replicate(100, experiment(epsilon = 0.005)))
```

Without data point

```{r}
experiment <- function(){
  rdtDP <- DPRandomDecisionTreeClassifier(Y = syntheticData[-1,3], X = syntheticData[-1,1:2], height = 2, numTrees = 1, epsilon = 0.005)
  predict(rdtDP, syntheticData[1,1:2])
}
table(replicate(100, experiment()))
```

So the two models become more "alike" and the adversary can't guess which dataset is actually used for each output.

The generated decision tree is below.
```{r}
#dtDP1$trees[[1]]$printTree()
#rdtDP2$trees[[1]]$printTree()
```

## Contributions

TODO

## Future extensions

So far, our package deals specifically with privatizing training data in the machine learning workflow. Oftentimes, machine learning algorithms have yet another set of data (the hold out set or validation set) for hyperparameter tuning and cross validations. These data also influence the final model output and so requires another set of privatization procedure to enforce differential privacy.

Again, **diffPrivML** serves as a precursor in standardizing a library of ML algorithms and workflows that have differential privacy baked in. Classification and output perturbations are just starting points. The package can naturally be extended to unsupervised learning as well as other perturbation methods (input and objective perturbations).

Thank you!

## References

Zhanglong Ji, Zachary C Lipton, and Charles Elkan. “Differential privacy and machine learning:
  A survey and review”. In: arXiv preprint arXiv:1412.7584 (2014).
  
Jaideep Vaidya, Basit Shafiq, Anirban Basu, and Yuan Hong. "Differentially
  private naive Bayes classification." In Web Intelligence, pages 571–576, 2013.
  
Geetha Jagannathan, Krishnan Pillaipakkamnatt, and Rebecca N. Wright.
  "A practical differentially private random decision tree classifier". In International
  Conference on Data Mining Workshops, pages 114–121, 2009.


