---
title: "Report-diffPrivML"
author: "Terence Tam, Yifan Gong"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Description

Give some background info on $\epsilon$-differential private mechanism. 

what does this package do? 
-precursor to standard mechanism with DF

## Use cases

Our use cases are mainly supported for classification tasks.

### Random Decision Tree Ensembles Classifier

First, let's invoke the library.

```{r}
library(diffPrivML)
```

Then, let's use a dataset preloaded in the diffPrivML package.
```{r}
data(BreastCancer)
head(BreastCancer)
```

### Naive Bayes Classifier

```{r}
library(caret)
data(iris)
y <- iris[, 5]
x <- iris[, 1:4]
trainingIndices <-createDataPartition(y, p = 0.70, list = FALSE)
trainingSetX <- x[trainingIndices, ]
trainingSetY <- y[trainingIndices]
testSetX <- x[-trainingIndices, ]
testSetY <- y[-trainingIndices]
naiveBayesDP <- DPNaiveBayesClassifier(trainingSetY, trainingSetX)
predictions <- predict(naiveBayesDP, testSetX)
accuracy <- sum(predictions == testSetY) / length(testSetY)
accuracy
```

Add the ggplot for accuracy vs epsilon

## Contributions


